---
title: "CMS025: Impact of a daily oral probiotic combination and antibiotic exposure on the gut microbiome of UK care home residents: secondary analysis of a randomised placebo-controlled trial"
output: html_document
date: "2023-10-04"
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
options(bitmapType = "cairo")
```

```{r load libraries, echo=FALSE, message=FALSE}

library(ggplot2)
library(RSQLite)
#library(dplyr)
library(phyloseq)
#library(dplyr)
library(OCMSutility)
library(Maaslin2)
library(propr)
library(NetCoMi)

```


# Overview

This analysis is concerned with doing differential abundance testing using Maaslin2 - we are always plagued with zeros in these types of analyses. I suspect it won't be much different to the results of DESeq2 but for visualisation the clr is often a nicer way to present as the zeros are imputed.

```{r read data, echo=FALSE, message=FALSE}

# read asv counts table
asv <- read.csv("~/work/CMS025/export/CMS025/data/taxa_abundances.tsv",
                header = TRUE,
                stringsAsFactors = FALSE,
                row.names=1,
                sep = "\t")

# remove the filtered and trimmed bits and bobs
colnames(asv) <- gsub("filtered\\.trimmed\\.", "", colnames(asv))

# read metadata
metadata <- read.csv("../../data/metadata/PRINCESS_metadata.txt",
                     header = TRUE,
                     stringsAsFactors = FALSE,
                     sep="\t",
                     row.names = 1)

# remove filtered and trimmed bits from rownames
rownames(metadata) <- sub("filtered-trimmed-", "", rownames(metadata))

# check colnames and rownames match and
# re-order
if(!(setequal(rownames(metadata), colnames(asv)))){
  metadata <- metadata[colnames(asv),]
}


# The community standard is still in here so take this out
metadata <- metadata[rownames(metadata) != "ZymobiomicsDNAStandard",]
asv <- asv[,colnames(asv) != "ZymobiomicsDNAStandard"]


# There are missing patient ids so presumably these need to go
metadata <- metadata[!(is.na(metadata$New_PID)),]
asv <- asv[,rownames(metadata)]


```

```{r get genus level data, echo=FALSE, message=FALSE}

# aggregate counts at the genus level
genera <- gsub("ASV[0-9].*:", "", rownames(asv))
genera <- gsub(";s__.*", "", genera)

genus_counts <- aggregate(asv, by=list(as.factor(genera)), FUN="sum")
rownames(genus_counts) <- genus_counts$Group.1
genus_counts <- genus_counts[,2:ncol(genus_counts)]

# Remove genera that are NA at genus level
genus_counts <- genus_counts[grep("g__NA", rownames(genus_counts), invert=TRUE),]

```


### Baseline Maaslin2

This analysis is to see whether we can detect differences between allocation groups at baseline. There certainly shouldn't be any differences but if there are they could impact on downstream interpretation.

```{r baseline deseq, echo=FALSE, message=FALSE}

meta_baseline <- metadata[metadata$Timepoint == "Baseline",]
genus_counts_baseline <- genus_counts[,rownames(meta_baseline)]

genera_to_keep <- rownames(genus_counts_baseline[rowSums(genus_counts_baseline > 0.01) > 25,])

maaslin_baseline <- Maaslin2(genus_counts_baseline,
                    meta_baseline,
                    "maaslin_out",
                    normalization="none",
                    fixed_effects = "Allocationcode",
                    reference="placebo")

maaslin_res <- maaslin_baseline$results


```

## Associations with frailty

```{r frailty, echo=FALSE, message=FALSE}

meta_baseline$frailty <- gsub(" ", ".", meta_baseline$FRAI_BL)
maaslin_baseline_frail <- Maaslin2(genus_counts_baseline,
                    meta_baseline,
                    "maaslin_out",
                    normalization="none",
                    fixed_effects = "frailty",
                    reference="frailty,Very.fit")

maaslin_res_frail <- maaslin_baseline_frail$results
maaslin_res_frail <- maaslin_res_frail[order(maaslin_res_frail$qval, decreasing=FALSE),]
```


## Associations with PPI

```{r PPI, echo=FALSE, message=FALSE}

maaslin_baseline_ppi <- Maaslin2(genus_counts_baseline,
                    meta_baseline,
                    "maaslin_out",
                    normalization="none",
                    fixed_effects = "PPI_BL",
                    reference="PPL_BL,No")

maaslin_res_ppi <- maaslin_baseline_ppi$results
maaslin_res_ppi <- maaslin_res_ppi[order(maaslin_res_ppi$qval, decreasing=FALSE),]
```


## Associations with Abx exposure {.tabset .tabset-pills}

Here I am splitting the analysis into two - one looking at the placebo group and the other looking at the probiotic group. This is based on observations of alterations in alpha diversity specifically in the probiotic group but not in the placebo group.

### Differential abundance ~ ABX exposure

```{r split data into placebo and probiotic, echo=FALSE, message=FALSE}

# split by placebo and probiotic (only looking at three months)
meta_placebo <- metadata[metadata$Timepoint == "Three-months" & metadata$Allocationcode == "placebo",]
meta_probiotic <- metadata[metadata$Timepoint == "Three-months" & metadata$Allocationcode == "probiotic",]

genus_counts <- genus_counts[,c(rownames(meta_placebo), rownames(meta_probiotic))]
genus_counts <- genus_counts[rowSums(genus_counts > 0) > ncol(genus_counts)/4,]

genus_placebo <- genus_counts[, rownames(meta_placebo)]
genus_probiotic <- genus_counts[, rownames(meta_probiotic)]


placebo_colors <- RColorBrewer::brewer.pal(5, "BuPu")
placebo_colors <- c("Exposed to antibiotics" = placebo_colors[5],
                    "Not exposed to antibiotics" = placebo_colors[2])

probiotic_colors <- RColorBrewer::brewer.pal(5, "Reds")
probiotic_colors <- c("Exposed to antibiotics" = probiotic_colors[5],
                    "Not exposed to antibiotics" = probiotic_colors[2])


maaslin_placebo <- Maaslin2(relab(genus_placebo),
                    meta_placebo,
                    "maaslin_out",
                    normalization="none",
                    fixed_effects = "ABXEXPOSED_3M",
                    reference="Not exposed to antibiotics")

maaslin_probiotic <- Maaslin2(relab(genus_probiotic),
                    meta_probiotic,
                    "maaslin_out",
                    normalization="none",
                    fixed_effects = "ABXEXPOSED_3M",
                    reference="Not exposed to antibiotics")


placebo_maaslin_res <- maaslin_placebo$results
probiotic_maaslin_res <- maaslin_probiotic$results

write.table(placebo_maaslin_res, file="export.dir/maaslin_placebo.tsv",
            row.names=FALSE,
            quote=FALSE,
            sep="\t")

write.table(probiotic_maaslin_res, file="export.dir/maaslin_probiotic.tsv",
            row.names=FALSE,
            quote=FALSE,
            sep="\t")

maaslin_sig_placebo <- maaslin_placebo$results[maaslin_placebo$results$qval < 0.25,]
maaslin_sig_probiotic <- maaslin_probiotic$results[maaslin_probiotic$results$pval < 0.05,]


```

## Maaslin interaction analysis

```{r maaslin interactions, echo=FALSE, message=FALSE}

genus_counts_int <- genus_counts[genera_to_keep, c(rownames(meta_placebo), rownames(meta_probiotic))]
coldata <- as.data.frame(dplyr::bind_rows(meta_placebo, meta_probiotic))


coldata_int <- model.matrix(~Allocationcode*ABXEXPOSED_3M, data = coldata)

coldata$interaction <- coldata_int[,"Allocationcodeprobiotic:ABXEXPOSED_3MNot exposed to antibiotics"]

maaslin_interaction <- Maaslin2(relab(genus_counts_int),
                    coldata,
                    "maaslin_out",
                    normalization="none",
                    fixed_effects = "interaction")


head(maaslin_interaction$results)
```





